---
title: "DATA 624 HW2 Test"
author: "Samuel I Kigamba"
date: "July 6, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 7.2. Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data:
# y = 10 sin(πx1x2) + 20(x3 − 0.5)2 + 10x4 + 5x5 + N(0, σ2) where the x values are random variables uniformly distributed between [0, 1] (there are also 5 other non-informative variables also created in the simulation). 
#The package mlbench contains a function called mlbench.friedman1 that simulates these data:

# Which models appear to give the best performance? Does MARS select the informative predictors (those named X1–X5)?


## Tune several models

```{r, message=FALSE, warning=FALSE}
library(caret)
library(dplyr)
library(tidyr)
library(corrplot)
library(AppliedPredictiveModeling)
library(e1071)
library(kernlab)
library(earth)
library(dplyr)
library(mlbench)
library(RANN)

```

```{r}

set.seed(200)

trainingData <- mlbench.friedman1(200, sd = 1)   ##We convert the 'x' data from a matrix to a data frame  > ##One reason is that this will give the columns names.  
trainingData$x <- data.frame(trainingData$x)  ##Look at the data using  > featurePlot(trainingData$x, trainingData$y)  > ##or other methods.  >  > ##This creates a list with a vector 'y' and a matrix  > ##of predictors 'x'. Also simulate a large test set to  > ##estimate the true error rate with good precision:  > 
testData <- mlbench.friedman1(5000, sd = 1)  
testData$x <- data.frame(testData$x)

```


KNN
```{r}

knnModel <- train(x = trainingData$x,  y = trainingData$y, method = "knn", preProc = c("center", "scale"),  tuneLength = 10)    

knnModel

```

```{r}
knnPred <- predict(knnModel, newdata = testData$x)   ##The function 'postResample' can be used to get the test set  
##perforamnce values  
postResample(pred = knnPred, obs = testData$y) 

```

MARS

```{r}
marsFit <- earth(trainingData$x, trainingData$y)
marsFit

```

```{r}
summary(marsFit)

```


Tune the model using external resampling
```{r}
marsGrid <- expand.grid(.degree = 1:2, .nprune = 2:38) 


marsFittune <- train(trainingData$x, trainingData$y, method = "earth", tuneGrid = marsGrid,  trControl = trainControl(method = "cv")) 
summary(marsFittune)

```


The optimal MARS model minimized the RMSE when the nprune = 13 and the degree = 2

```{r}
head(predict(marsFittune, testData$x))

```

Look at importance of predictors

```{r}
varImp(marsFittune) 

```

MARS selected x1 as the most important predictor and x4 as second.

SVM

```{r}
svmRTuned <- train(trainingData$x, trainingData$y,   method = "svmRadial",   preProc = c("center", "scale"),   tuneLength = 14,   trControl = trainControl(method = "cv")) 

svmRTuned

```

```{r}
svmRTuned$finalModel

```


The model used 152 training set data points as support vectors. The MARS Model fit best with a high R2 of ~.95



## 7.5. Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.

# (a) Which nonlinear regression model gives the optimal resampling and test set performance?


```{r}
data(ChemicalManufacturingProcess)

set.seed(56)

knnmodel2 <- preProcess(ChemicalManufacturingProcess, "knnImpute")
df <- predict(knnmodel2, ChemicalManufacturingProcess)

df <- df %>%
  select_at(vars(-one_of(nearZeroVar(., names = TRUE))))

in_train <- createDataPartition(df$Yield, times = 1, p = 0.8, list = FALSE)
train_df <- df[in_train, ]
test_df <- df[-in_train, ]

```

```{r}
knn_model <- train(
  Yield ~ ., data = train_df, method = "knn",
  center = TRUE,
  scale = TRUE,
  trControl = trainControl("cv", number = 10),
  tuneLength = 25
)
knn_model

```

```{r}
knn_predictions <- predict(knn_model, test_df)

results <- data.frame(t(postResample(pred = knn_predictions, obs = test_df$Yield))) %>%
  mutate("Model"= "KNN") 
results

```


MARS

```{r}
MARS_grid <- expand.grid(.degree = 1:2, .nprune = 2:38)

MARS_model <- train(
  Yield ~ ., data = train_df, method = "earth",
  tuneGrid = MARS_grid,
  # If the following lines are uncommented, it throws an error
  #center = TRUE,
  #scale = TRUE,
  trControl = trainControl("cv", number = 10),
  tuneLength = 25
)
MARS_model

```


The optimal MARS model minimized the RMSE when the nprune = 13 and the degree = 2

```{r}
head(predict(MARS_model, test_df))

```

SVM

```{r}
SVM_model <- train(
  Yield ~ ., data = train_df, method = "svmRadial",
  center = TRUE,
  scale = TRUE,
  trControl = trainControl(method = "cv"),
  tuneLength = 25
)
SVM_model

```


```{r}
SVM_predictions <- predict(SVM_model, test_df)

results <- data.frame(t(postResample(pred = SVM_predictions, obs = test_df$Yield))) %>%
  mutate("Model"= "SVM")

```


The SVM model was the best model according to the R2

# (b) Which predictors are most important in the optimal nonlinear regression model? Do either the biological or process variables dominate the list? How do the top ten important predictors compare to the top ten predictors from the optimal linear model?


```{r}
varImp(SVM_model, 10)

```


The processing (manufacturing) variables are most important with mfgprocess13 as most important.



# (c) Explore the relationships between the top predictors and the response for the predictors that are unique to the optimal nonlinear regression model. Do these plots reveal intuition about the biological or process predictors and their relationship with yield?


```{r}
df %>% select(c('ManufacturingProcess32','ManufacturingProcess13','BiologicalMaterial06','ManufacturingProcess17','BiologicalMaterial03','Yield')) %>% cor() %>% corrplot(method = 'circle')

```

Mfgprocess32 shows to have a high positive correlation, while Mfg Process 13 has a strong negative correlation. Biological Material 03 and 06 have positive correlations.













